---
title: Principal component analysis
author: Miguel Rodo
date: "2025-02-26"
bibliography: zotero.bib
format:
  html:
    embed-resources: true
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
nocite: |
    @johnson_wichern07
---

```{r }
#| include: false
library(tibble)
library(ggplot2)
for (x in list.files("R", full.names = TRUE)) {
  source(x)
}
```

# Context

- Multivariate analysis concerns situations where, for one reason or another, the number of variables causes some difficulty.
- **Example**: Meteorologists analyze temperature, humidity, wind speed, and more to understand weather systems.
  - Possible questions:
    - What sets of variable(s) are correlated?
    - Can we predict the likelihood of rain based on these variables?
    - What variables most distinguish between different weather patterns?
- For example:
  - **Exploratory data analysis**:
    - If you have a few variables, you can do all pair-wise scatterplots. But with 1000 variables, this is impractical.
  - **Identifying sets of correlated variables**:
    - You could do a heatmap of the correlations between variables. But the heatmap is very difficult to interpret with 1000 variables, so we would either needto organise the heatmap more effectively or another way to summarise the data.
  - **Too many explanatory variables**:
    - If we have more explanatory variables than observations, we cannot fit the model
    - The model may take too long to fit
    - The model may overfit
    - The model may be too complex to interpret
  - **Correlated explanatory variables**:
    - Correlated explanatory variables have more variable estimated effects. This can be a problem in small datasets, but large datasets are just more likely to have correlated variables.
  - **Multiple dependent variables**:
    - We may more accurately fit regression models with multiple dependent variables. This makes the regression analysis hopefully more accurate and robust, but complicates the structure and requires different methods.

# Basis of approach

- **Linear combinations**:

$$
a_1 x_1 + a_2 x_2 + \ldots + a_p x_p
$$ 

- Why linear combinations?
  - Simplicity >> strong theoretical results >> robustness, speed and interpretability

# Types of multivariate analysis

- $\symbf{X} \sim \symbf{X}$: PCA, factor analysis, correspondence analysis (count data)
  - **Examples**: identify correlated sets of variables
- $\symbf{Y} \sim \symbf{X}$: Multiple/multivariate regression, canonical correlation analysis, discrimination and classification

# Principal components analysis

- So, the first thing we're going to consider is the construction of new variables.
- Now, we're going to restrict ourselves to linear combinations of the original variables.
- Why?
  - Well, it's got an algebraic solution that's fast to fit, they're more interpretable, and they're robust. 
- Now, how do we choose the linear combinations?
  - So, let's just think about the first variable. 
- So, we'll suppose that each observation is $\symbf{X}\in\mathcal{R}^p$, with zero mean and covariance matrix $\symbf{\Sigma}$.
  - Note that we don't assume that the variables are independent, or that they're normally distributed.
- Now, we have $p$ variables, and we want ot get "better" variables such that we need fewer of them.
- So, let's think about getting the first variable, let's call it $Y_1$. Let's constrain $Y_1$ to be a linear combination of the original variables, so $Y_1 = \symbf{X}^T\symbf{a}_1$:

$$
\begin{align*}
\symbf{Y}_1 &= a^{(1)}_1X_1 + a^{(1)}_2X_2 + \ldots + a^{(1)}_pX_p, \\
&= \symbf{a}^{(1)'}\symbf{X}. \\
\end{align*}
$$

- Okay, so, now, we need to decide how to choose $\symbf{a}_1$.
- Well, what we're going to do is choose $\symbf{a}_1$ such that the variance of $Y_1$ is maximised.

Below is a revised version with more bullet points and shorter sentences, ideal for an HTML page walk-through:

### Why maximise variance?

- **Consider the alternative:**  
  - Imagine choosing a linear combination that minimizes variance.
  - If one variable has zero variance, putting all the weight on it means every observation is the same.
  - Such a variable provides no useful information in prediction or classification.

- **Low variance = low information:**  
  - Even if a variable doesn’t have zero variance, very low variance means little change between observations.
  - For example, if height varies only from 1.79 m to 1.81 m, it is less informative.
  - In contrast, if weight ranges from 50 kg to 100 kg, it can capture more meaningful differences.

- **Statistical benefits:**  
  - In linear regression, the variance of a regression coefficient is inversely related to the predictor’s variance.
  - Higher variance in a variable leads to lower standard errors and more precise estimates.

- **Leveraging covariance:**  
  - The variance of a linear combination is given by:  
    $$
    \mathrm{Var}(a_1X_1 + a_2X_2) = a_1^2\mathrm{Var}(X_1) + a_2^2\mathrm{Var}(X_2) + 2a_1a_2\mathrm{Cov}(X_1, X_2)
    $$
  - If two variables are positively correlated, using positive weights increases the overall variance.
  - PCA takes advantage of this by upweighting correlated variables.

- **Efficient information capture:**  
  - Maximising variance helps identify a reduced set of variables that captures the most information.
  - Specifically, this means you can reduce the data from $p$ variables to a smaller number $r$ (where $r < p$) such that we can reconstruct the original $p$ variables with minimal least squares reconstruction error.

- **Joint display of variables and observations:**:
  - Standard scatterplots are limited to two variables.
  - PCA allows us to visualize the joint display of all variables and observations in a lower-dimensional space, such that we can read (approximately) the values of the original variables for each observation all in one plot.

- **Key takeaways:**  
  - **Selects informative directions:** Captures meaningful differences among observations.
  - **Enhances stability:** Leads to more precise estimates in modeling.
  - **Reflects data structure:** Emphasizes variables that are jointly informative.
  - **Enables efficient compression:** Supports dimensionality reduction with minimal reconstruction error.
  - **Joint display:** Allows visualization of all variables and observations in a lower-dimensional space.

## Objective function for maximising variance

- Okay, so the objective function now looks like this:

$$
\symbf{a}^{(1)} = \underset{\symbf{a}^{(1)}}{\arg\max} \, \mathrm{Var}(\symbf{Y}_1) = \underset{\symbf{a}^{(1)}}{\arg\max} \, \symbf{a}^{(1)'}\symbf{\Sigma}\symbf{a}^{(1)}.
$$

- What's the problem?
  - Well, consider the scenario where $\symbf{X} = \begin{pmatrix} X_1 & X_2 \end{pmatrix}'$, and $\symbf{\Sigma}=\symbf{I}_2$.
  - In this case, $Y_1 = a_1X_1 + a_2X_2$ and $VAR(Y_1) = a_1^2 + a_2^2$. So, we can make $VAR(Y_1)$ arbitrarily large by making $a_1$ and $a_2$ large.

- So, we restrict the solution to $\symbf{a}^{(1)'}\symbf{a}^{(1)} = 1$, i.e. $\symbf{a}^{(1)}$ is a unit vector.

- We aim to maximize the variance of $Y_1$, which is a linear combination of the original variables, with respect to $\symbf{a}_1$. The objective function is:

$$
\mathrm{Var}(Y_1) = \symbf{a}_1'\symbf{\Sigma}\symbf{a}_1.
$$

- To achieve this, we solve the following optimization problem:

$$
\underset{\symbf{a}_1}{\max} \, \mathrm{Var}(Y_1) = \underset{\symbf{a}_1}{\max} \, \symbf{a}_1'\symbf{\Sigma}\symbf{a}_1,
$$

subject to $\symbf{a}_1'\symbf{a}_1 = 1$.

---

### Objective function for the second principal component

- **Why a second principal component?**
  - Typically, one principal component isn't enough—we often want multiple variables to summarize the data.
  - But we can't just maximise variance again without restrictions—this would produce the same linear combination as before.

- **How do we fix this?**
  - We add a constraint that the second principal component must be **uncorrelated** with the first.
  - Two linear combinations are uncorrelated if their coefficient vectors ($\symbf{a}_1$ and $\symbf{a}_2$) are **orthogonal** (dot product equals zero).

- **New optimisation problem:**
  - Maximise the variance of $Y_2$:
  $$
  Y_2 = \symbf{X}'\symbf{a}_2
  $$
  subject to two constraints:
  1. Unit length:
  $$
  \symbf{a}_2'\symbf{a}_2 = 1
  $$
  2. Orthogonality with the first principal component:
  $$
  \symbf{a}_2'\symbf{a}_1 = 0
  $$
  
- **Optimisation function:**
  $$
  \symbf{a}_2 = \underset{\symbf{a}_2}{\arg\max}\,\symbf{a}_2'\symbf{\Sigma}\symbf{a}_2
  $$
  subject to:
  $$
  \symbf{a}_2'\symbf{a}_2 = 1,\quad\text{and}\quad \symbf{a}_2'\symbf{a}_1 = 0
  $$

---

### Objective function for the $k$-th principal component

- **Generalising further:**
  - For the $k$-th principal component, we continue the same logic.
  - It must maximise variance but remain uncorrelated (orthogonal) to **all previous principal components**.

- **Optimisation problem:**
  - Maximise the variance of $Y_k$:
  $$
  Y_k = \symbf{X}'\symbf{a}_k
  $$
  subject to:
  1. Unit length constraint:
  $$
  \symbf{a}_k'\symbf{a}_k = 1
  $$
  2. Orthogonality constraints:
  $$
  \symbf{a}_k'\symbf{a}_j = 0,\quad\text{for all } j<k
  $$

- **Optimisation function:**
  $$
  \symbf{a}_k = \underset{\symbf{a}_k}{\arg\max}\,\symbf{a}_k'\symbf{\Sigma}\symbf{a}_k
  $$
  subject to:
  $$
  \symbf{a}_k'\symbf{a}_k = 1,\quad\text{and}\quad\symbf{a}_k'\symbf{a}_j = 0,\;\text{for all }j < k
  $$

---

### How many principal components can we have?

- Because the variance-covariance matrix $\symbf{\Sigma}$ is at least positive semi-definite:
  - You can obtain **at most** as many principal components as you have original variables (i.e., $p$).
  - Each additional principal component captures less variance than the one before.
  
- Eventually, later principal components may have zero variance:
  - This occurs when the eigenvalues of the covariance matrix become zero.
  - Such components provide no additional information beyond earlier ones and indicate redundancy in the data.

- Practically, this tells us how many principal components we really need:
  - Typically, you choose the first few principal components that capture most of the variance.
  - Additional components add little useful information and may only represent noise.

# Example applications

- **Market analysis**: Identify correlated stocks to ensure a diversified portfolio
- **Image compression**: Represent images with fewer pixels while preserving accuracy
- **Bioinformatics**: Identify sets of co-expressed genes responsible for different functions in the body


# How do we find the principal components?

- Up until this point, we've explored various options for $\symbf{a}_k$ without any method. We'll now discuss how to compute these principal components.
- First, we'll discuss a more general result.

# Theorem 1: Maximisation of quadratic forms for points on the unit sphere

Let $\symbf{B}:p\times p$ be a positive semi-definite matrix with the $i$-th largest eigenvalue $\lambda_i$ and associated eigenvector $\symbf{e}_i$. Then we have that 

$$
\max_{\symbf{x}\neq \symbf{0}} \frac{\symbf{x}'\symbf{B}\symbf{x}}{\symbf{x}'\symbf{x}}=\lambda_1 \; (\mathrm{attained}\; \mathrm{when} \; \symbf{x}=\symbf{e}_1)
$$

and that

$$
\max_{\symbf{x}\neq \symbf{0},\symbf{x}\perp \symbf{e}_1, \symbf{e}_2, \ldots, \symbf{e}_{i-1}} \frac{\symbf{x}'\symbf{B}\symbf{x}}{\symbf{x}'\symbf{x}}=\lambda_i \; (\mathrm{attained}\; \mathrm{when} \; \symbf{x}=\symbf{e}_{i})
$$

for $i\in\{2,3, \ldots, p\}$, where $\lambda_i$ is the $i$-th largest eigenvalue and $\symbf{e}_i$ is the associated eigenvector.

# Proof of Theorem 1 {.smaller}

Let $\symbf{P}:p\times p$ be the orthogonal matrix whose $i$-th column is the $i$-th eigenvector and $\Lambda$ be the diagonal matrix with ordered eigenvalues along the diagonal. Let $\symbf{B}^{1/2}=\symbf{P}\symbf{\Lambda}^{1/2}\symbf{P}'$ and $\symbf{y}=\symbf{P}'\symbf{x}$.

First, we show that the quadratic form can never be larger than $\lambda_1$:

\begin{align*} 
\frac{\symbf{x}'\symbf{B}\symbf{x}}{\symbf{x}'\symbf{x}} &=  \frac{\symbf{x}'\symbf{B}^{1/2}\symbf{B}^{1/2}\symbf{x}}{\symbf{x}'\symbf{P}\symbf{P}'\symbf{x}} 
= \frac{\symbf{x}'\symbf{P}\symbf{\Lambda}^{1/2}\symbf{P}'\symbf{P}\symbf{\Lambda}^{1/2}\symbf{P}'\symbf{x}}{\symbf{y}'\symbf{y}}
=  \frac{\symbf{y}'\symbf{\Lambda}\symbf{y}}{\symbf{y}'\symbf{y}} \\ 
 &=  \frac{\sum_{i=1}^p\lambda_iy_i^2}{\sum_{i=1}^py_i^2} 
 \leq \lambda_1 \frac{\sum_{i=1}^py_i^2}{\sum_{i=1}^py_i^2}
 = \lambda_1 
\end{align*}

# Proof of Theorem 1 (cont.) {.smaller}

Now we show that this is actually attained for $\symbf{x}=\symbf{e}_1$. Since eigenvectors are by convention length 1, we consider $\symbf{e}_1'\symbf{B}\symbf{e}_1$.

First, let $\symbf{c}_i$ be the unit vector with a 1 in the $i$-th position (and 0's everywhere else).

Expanding $\symbf{B}$ by the eigen decomposition, we have that

$$
\symbf{e}_1'\symbf{B}\symbf{e}_1=\symbf{e}_1'\symbf{P}\symbf{\Lambda}\symbf{P}'\symbf{e}_1.
$$

Since $B$ is symmetric, the eigenvectors can be chosen orthogonal.
We do so, which implies that $\symbf{e}_i'P=\symbf{c}_i'$, where $\symbf{c}_i$ is the unit vector with a 1 in the $i$-th position (and 0 everwhere else).
Consequently,

$$
\symbf{e}_i'\symbf{P}\symbf{\Lambda}\symbf{P}'\symbf{e}_i
= \symbf{c}_i'\symbf{\Lambda}\symbf{c}_i=\lambda_i,
$$

and so $\symbf{e}_1'\symbf{P}\symbf{\Lambda}\symbf{P}'\symbf{e}_1=\lambda_1$.

# Proof of Theorem 1 (cont.) {.smaller}

Now, we consider the case where $\symbf{x}$ is orthogonal to $\symbf{e}_1, \symbf{e}_2, \ldots, \symbf{e}_{i-1}$.

Each component in the vector $\symbf{y}$ is the dot product of $x$ and an eigenvector $\symbf{e}_i$.
Since we choose $x$ orthogonal to the first $i-1$ eigenvectors, the first $i-1$ entries of $\symbf{y}$ are zero.

Returning to considering the quadratic form, we have that

\begin{align}
\frac{\symbf{x}'\symbf{B}\symbf{x}}{\symbf{x}'\symbf{x}} 
&= \frac{\sum_{j=1}^p\lambda_jy_j^2}{\sum_{j=1}^py_j^2} \\
&= \frac{\sum_{j=i}^p\lambda_jy_j^2}{\sum_{j=i}^py_j^2} 
 \leq \lambda_i \frac{\sum_{j=i}^py_j^2}{\sum_{j=i}^py_j^2} 
 = \lambda_i
\end{align}

Using the same argument as before, we can show that this is actually attained for $\symbf{x}=\symbf{e}_i$.

# Connection with principal components

- Theorem 1 has already done the heavy lifting, as we can simply set $\symbf{B}=\symbf{\Sigma}$.
  - To see this, consider that $\frac{\symbf{x}'\symbf{\Sigma}\symbf{x}}{\symbf{x}'\symbf{x}}=\frac{\symbf{x}'\symbf{\Sigma}\symbf{x}}{||\symbf{x}||^2} = (\symbf{x}/||\symbf{x}||)'\symbf{\Sigma}(\symbf{x}/||\symbf{x}||)$, so implicitly the constraint $||\symbf{x}||=1$ is already present.
- All that's left to do is place Theorem 1 in a probabilistic context, and relate it to the variance and covariance of linear combinations of a random vector.

# Theorem 2: Selecting principal components {.smaller}

Let $\symbf{X}:\symbf{p}\times 1$ be the random vector with variance-covariance matirx $\symbf{\Sigma}:p\times p$, which has the $i$-th largest eigenvalue as $\lambda_i$ and associated eigenvector $\symbf{e}_i$.

Then the $i$-th principal component (defined as before) is given by 

$$
Y_i = \symbf{e}_i'\symbf{X},
$$

implying that

\begin{align*}
\mathrm{Var}[Y_i] &= \lambda_i \;\forall \; i \in \{1,2, \ldots, p\}, \; \mathrm{and} \\
\mathrm{Cov}[Y_i, Y_j] &= 0 \; \mathrm{for} \; i\neq j.
\end{align*}

# Proof of Theorem 2 {.smaller}

From Theorem 1, we know that 

$$
\max_{\symbf{x}} \frac{\symbf{x}'\symbf{\Sigma}\symbf{x}}{\symbf{x}'\symbf{x}} = \lambda_1,
$$

which we achieve by setting $\symbf{x}=\symbf{e}_1$.

Since $||\symbf{e}_1||=1$, $\frac{\symbf{e}_1'\symbf{\Sigma}\symbf{e}_1}{\symbf{e}_1'\symbf{e}_1}=\symbf{e}_1'\symbf{\Sigma}\symbf{e}_1$, which is equal to $\mathrm{Var}[\symbf{a}'\mathrm{X}]=\mathrm{Var}[Y_1]$ if we set $\symbf{a}=\symbf{e}_1$. This implies that $\symbf{a}=\symbf{e}_1$ maximises the variance, which is $\lambda_1$.

Analagous reasoning shows that $\mathrm{Var}[\symbf{a}_i'\symbf{X}]$ has a maximum value $\lambda_i$ for $i\in\{2,3,\ldots,p\}$ that is attained when we set $\symbf{a}_i=\symbf{e}_i$, under the restriction $\symbf{a}_i$ to be orthogonal to $\symbf{a}_1, \symbf{a}_2, \ldots, \symbf{a}_{i-1}$.

For $i\neq j$, $\mathrm{Cov}[\symbf{e}_i'\symbf{X},\symbf{e}_j'\symbf{X}] = \symbf{e}_i'\symbf{\Sigma}\symbf{e}_j = 0$.

# A note on terminology

- Principal components (the $\symbf{Y}_i$'s) may also be referred to as "scores".
- The coefficient vectors $\symbf{a}_i$ may also be referred to as "loadings" (thinking algebraically) or as "principal axes"/"principal directions" (thinking geometrically).

# Example {.smaller}

::::{.columns}

:::{.column width="50%"}

Suppose that we have the following variance-covariance matrix:

```{r}
#| echo: true
vcov_mat <- matrix(
  c(1, 0.9, 0.9, 0, 0, 0, 0,
    0.9, 1, 0.9, 0, 0, 0, 0,
    0.9, 0.9, 1, 0, 0, 0, 0,
    0, 0, 0, 1, 0.9, 0.9, 0.9,
    0, 0, 0, 0.9, 1, 0.9, 0.9,
    0, 0, 0, 0.9, 0.9, 1, 0.9,
    0, 0, 0, 0.9, 0.9, 0.9, 1),
  byrow = TRUE,
  nrow = 7
)
colnames(vcov_mat) <- paste0("V", 1:7)
rownames(vcov_mat) <- paste0("V", 1:7)
```

:::

:::{.column width="50%"}

```{r}
#| results: asis
knitr::kable(vcov_mat)
```

:::

::::

# Example (cont.) {.smaller}

::::{.columns}

:::{.column width="50%"}

- We apply the eigen decomposition to the variance-covariance matrix:

```{r}
#| echo: true
eig_obj <- eigen(vcov_mat)
```

- to obtain the following eigenvectors:

```{r}
#| echo: true
eig_vec_mat <- eig_obj$vectors
rownames(eig_vec_mat) <- paste0("V", 1:7)
colnames(eig_vec_mat) <- paste0("PC", 1:7)
```

:::

:::{.column width="50%"}

```{r}
eig_vec_mat |> signif(2)
```

:::

::::

# Example (cont.) {.smaller}

::::{.columns}

:::{.column width="50%"}

- By the variance formula, these linear combinations have the following variances:

```{r}
#| echo: true
(eig_vec_mat |> t()) %*%
  vcov_mat %*%
  (eig_vec_mat) |>
  diag() |>
  signif(2)
```

:::

:::{.column width="50%"}

- These are the same as the eigenvalues (by theorem 2):

```{r}
#| echo: true
eig_obj$values |>
  signif(2) |>
  setNames(paste0("PC", 1:7))
```

- Clearly, due to the correlation structure, the orthogonality constraint implies that only two PCs have meaningful variance.

:::

::::

