---
title: Principal component analysis
author: Miguel Rodo
date: "2025-04-07"
bibliography: zotero.bib
format:
  html:
    embed-resources: true
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
nocite: |
    @johnson_wichern07
---

```{r }
#| include: false
library(tibble)
library(ggplot2)
fn_vec <- list.files("R", full.names = TRUE)
for (i in seq_along(fn_vec)) {
  source(fn_vec[i])
}
```

## Context

- Multivariate analysis addresses situations where many variables make analysis difficult.
- Example multivariate analysis: Meteorologists analyze temperature, humidity, wind speed, and other variables to understand weather patterns.
  - Typical questions:
    - Which variables are correlated?
    - Can we predict rainfall from these variables?
    - What variables distinguish clearly between weather systems?

### Common issues in multivariate analysis

- **Exploratory data analysis:**
  - With a few variables, pairwise scatterplots are easy.
  - With many variables (e.g. 1000), scatterplots become impractical.

- **Identifying correlated variables:**
  - Correlation heatmaps quickly become cluttered as variable count increases.
  - Effective interpretation requires summarisation methods or smarter organisation.

- **Excess explanatory variables:**
  - More explanatory variables than observations:
    - Models can't be fitted (identifiability issue).
    - Models may take too long or overfit.
    - Interpretation becomes overly complex.

- **Correlated explanatory variables:**
  - Highly correlated variables produce unstable parameter estimates.
  - Larger datasets increase the likelihood of correlated variables.

- **Multiple dependent variables:**
  - Multivariate regression may increase accuracy and robustness.
  - However, such models require specialised methods and interpretations.


## Types of multivariate analyses

- **Analysing relationships among explanatory variables ($\symbf{X}\sim\symbf{X}$):**
  - PCA, factor analysis, correspondence analysis (for count data)
  - Aim: identify groups of correlated variables.

- **Analysing relationships between dependent and explanatory variables ($\symbf{Y}\sim\symbf{X}$):**
  - Multivariate regression, canonical correlation analysis, discriminant analysis, classification.

### Unifying theme

The unifying theme of the techniques we'll discuss is *linearity*:

$$
a_1 x_1 + a_2 x_2 + \dots + a_p x_p
$$

- Linear combinations of variables are the basis of many multivariate methods. When people think "multivariate analysis", this is often what they mean.
- Linear techniques were traditionally the only ones available (due to computational constraints), but they are still widely used for the following reasons:
  - They are simple to understand and interpret.
  - They are fast to compute.
  - They suffer less from fragility due to excessive flexibility.
- It is also important to understand the foundational approaches, as they are often used as building blocks for more complex methods. More complex methods are also often discussed in relation to linear methods, so it is important to understand the linear methods first.

## Principal component analysis

- Principal component analysis (PCA) is a linear method for reducing the dimensionality of data.
- The easiest way to think about PCA is that we want to create a smaller number of new variables that have as much information as possible about the original variables.

### Choosing the linear combinations

- Consider a random vector $\symbf{X}\in\mathbb{R}^p$ with zero mean and covariance matrix $\symbf{\Sigma}$.
- Goal: reduce $p$ variables to fewer variables while keeping most information.
- First principal component ($Y_1$) is a linear combination of original variables:

$$
Y_1 = \symbf{a}_1'\symbf{X} = a_{1}^{(1)}X_1 + a_{2}^{(1)}X_2 + \dots + a_{p}^{(1)}X_p
$$

- How do we select $\symbf{a}_1$?  
  - Choose $\symbf{a}_1$ to maximise the variance of $Y_1$.

## Why maximise variance?

### Minimising variance loses information

- Consider choosing a linear combination with minimal variance:
  - If a variable has zero variance, all observations are identical.
  - This provides no predictive or discriminative information.

### Low variance implies low information

- Variables with minimal variation are less informative:
  - Example: Height (1.79–1.81 m) vs. weight (50–100 kg).
  - Height provides little practical information compared to weight.

### Statistical benefits of high variance

- Regression coefficients' precision improves with higher variance predictors.
- Larger variance → lower standard errors → more precise estimation.

### Leveraging covariance structure

- The variance of linear combinations:
$$
\mathrm{Var}(a_1X_1 + a_2X_2) = a_1^2\mathrm{Var}(X_1) + a_2^2\mathrm{Var}(X_2) + 2a_1a_2\mathrm{Cov}(X_1,X_2)
$$
- Positively correlated variables given positive weights increase combined variance.
- PCA exploits correlations by assigning larger weights to correlated variables.

### Efficient representation of information

- PCA reduces dimensionality (from $p$ to $r<p$) with minimal information loss.
- Specifically, PCA minimises the reconstruction error (least squares sense).

### Joint display of variables and observations

- Standard scatterplots limit visualisation to two variables.
- Choosing the linear combinations to maximise variation leads to a low-dimensional plot displaying multiple variables simultaneously, aiding interpretation.

## Key advantages of PCA

- Selects informative directions capturing meaningful differences.
- Enhances statistical stability and precision.
- Reflects underlying correlation structure.
- Enables efficient data compression and visualisation.
- Minimises information loss during dimensionality reduction.

### Practical examples of PCA

- **Finance:** Reducing correlated stock returns for diversification.
- **Image compression:** Representing images with fewer, informative pixels.
- **Genomics:** Finding gene-expression patterns indicating biological processes.

## Maximising variance mathematically

Suppose that $\symbf{X}$ is a $p$-dimensional random vector with zero mean and covariance matrix $\symbf{\Sigma}$.
We want to find the linear combination of the variables that maximises the variance.

### Objective function for the first principal component

- The objective for choosing the first principal component:
$$
\symbf{a}_1 = \underset{\symbf{a}_1}{\arg\max}\; \symbf{a}_1'\symbf{\Sigma}\symbf{a}_1
$$
- Issue: Without constraints, this variance is unbounded.
- Solution: Restrict $\symbf{a}_1$ to be a unit vector ($\symbf{a}_1'\symbf{a}_1=1$).

### Height and weight example

- Suppose we have a dataset with two correlated variables, perhaps weight and height.
- Then, the scatterplot of these may look something like this:

```{r}
#| include: false
set.seed(2106)
plot_tbl <- tibble::tibble(
  height = rnorm(100, 160, 10)
) |>
  dplyr::mutate(
    weight = height / 2.5 + rnorm(100, 0, 3.5 * height / 160)
  ) |>
  dplyr::mutate(
    across(everything(), function(x) (x - mean(x)) / sd(x))
  ) |>
  dplyr::arrange(height)
hw_tbl <- plot_tbl
p <- ggplot(
  aes(x = height, y = weight),
  data = plot_tbl
) +
  geom_vline(
    xintercept = 0, col = "gray50", linetype = "solid"
  ) +
  geom_hline(
    yintercept = 0, col = "gray50", linetype = "solid"
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme( 
    panel.background = element_rect(fill = "white"), 
    plot.background = element_rect(fill = "white")
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(
    x = "Height",
    y = "Weight"
  ) +
  geom_point(alpha = 0.85, col = "dodgerblue")
path_p_height_weight <- projr::projr_path_get("cache", "fig", "p-height-weight.png")
cowplot::save_plot(
  plot = p,
  filename = path_p_height_weight,
  units = "cm",
  base_height = 5,
  base_width = 5
)#| echo: false
```

```{r}
#| echo: false
knitr::include_graphics(path_p_height_weight)
```

- The first PCA combines these two variables into one: $a_1 \times \text{height} + a_2 \times \text{weight}$.
- For example, suppose we have the point $[1.2, 0.3]$:
  - If we set $\symbf{a}' = [1/\sqrt{2}, 1/\sqrt{2}]$, then the new variable is $1/\sqrt{2} \times 1.2 + 1/\sqrt{2} \times 0.3$.
  - We go from the matrix of variables

```{r}
tibble::tibble(height = 1.2, weight = 0.3) |> as.data.frame()
```

- to

```{r}
tibble::tibble(size = 1.2 / sqrt(2) + 0.3 / sqrt(2)) |> as.data.frame()
```

- The first principal component can be viewed as the "size" of the observation, as height and weight contribute equally to the new variable and it increases with both height and weight.

## Selecting additional principal components

- One component typically insufficient.

### Objective for second principal component

- Maximising variance again without constraints yields identical results.
- Solution: Require second component ($\symbf{a}_2$) orthogonal to first ($\symbf{a}_1'\symbf{a}_2=0$):
$$
\symbf{a}_2 = \underset{\symbf{a}_2}{\arg\max}\; \symbf{a}_2'\symbf{\Sigma}\symbf{a}_2
$$
subject to:
$$
\symbf{a}_2'\symbf{a}_2=1,\quad \symbf{a}_2'\symbf{a}_1=0
$$

#### Height and weight data

- If the first principal component is $[1/\sqrt{2}, 1/\sqrt{2}]$, the second principal component has to be $[1/\sqrt{2}, -1/\sqrt{2}]$.
  - This captures the "shape" of the observation, as taller and thinner people have higher values, while shorter and wider people have lower values.
- For the example point of $[1.2, 0.3]$, the second principal component is $1/\sqrt{2} \times 1.2 - 1/\sqrt{2} \times 0.3= 0.64$.

### General objective for $k$-th principal component

- Generalise further: $k$-th component orthogonal to all previous components.
- Optimisation:
$$
\symbf{a}_k = \underset{\symbf{a}_k}{\arg\max}\; \symbf{a}_k'\symbf{\Sigma}\symbf{a}_k
$$
subject to:
$$
\symbf{a}_k'\symbf{a}_k=1,\quad \symbf{a}_k'\symbf{a}_j=0,\; \text{for all } j<k
$$

## How many principal components?

- At most, as many components as original variables ($p$).
- Later components capture progressively less variance.
- Components with zero variance (due to zero eigenvalues) indicate redundancy.
- Typically, select a small number of principal components that explain most variance.

## Theorem 1: Maximisation of quadratic forms for points on the unit sphere

Let $\symbf{B}:p\times p$ be a positive semi-definite matrix with the $i$-th largest eigenvalue $\lambda_i$ and associated eigenvector $\symbf{e}_i$. Then we have that 

$$
\max_{\symbf{x}\neq \symbf{0}} \frac{\symbf{x}'\symbf{B}\symbf{x}}{\symbf{x}'\symbf{x}}=\lambda_1 \; (\mathrm{attained}\; \mathrm{when} \; \symbf{x}=\symbf{e}_1)
$$

and that

$$
\max_{\symbf{x}\neq \symbf{0},\symbf{x}\perp \symbf{e}_1, \symbf{e}_2, \ldots, \symbf{e}_{i-1}} \frac{\symbf{x}'\symbf{B}\symbf{x}}{\symbf{x}'\symbf{x}}=\lambda_i \; (\mathrm{attained}\; \mathrm{when} \; \symbf{x}=\symbf{e}_{i})
$$

for $i\in\{2,3, \ldots, p\}$, where $\lambda_i$ is the $i$-th largest eigenvalue and $\symbf{e}_i$ is the associated eigenvector.

### Proof of Theorem 1 {.smaller}

Let $\symbf{P}:p\times p$ be the orthogonal matrix whose $i$-th column is the $i$-th eigenvector and $\Lambda$ be the diagonal matrix with ordered eigenvalues along the diagonal. Let $\symbf{B}^{1/2}=\symbf{P}\symbf{\Lambda}^{1/2}\symbf{P}'$ and $\symbf{y}=\symbf{P}'\symbf{x}$.

First, we show that the quadratic form can never be larger than $\lambda_1$:

\begin{align*} 
\frac{\symbf{x}'\symbf{B}\symbf{x}}{\symbf{x}'\symbf{x}} &=  \frac{\symbf{x}'\symbf{B}^{1/2}\symbf{B}^{1/2}\symbf{x}}{\symbf{x}'\symbf{P}\symbf{P}'\symbf{x}} 
= \frac{\symbf{x}'\symbf{P}\symbf{\Lambda}^{1/2}\symbf{P}'\symbf{P}\symbf{\Lambda}^{1/2}\symbf{P}'\symbf{x}}{\symbf{y}'\symbf{y}}
=  \frac{\symbf{y}'\symbf{\Lambda}\symbf{y}}{\symbf{y}'\symbf{y}} \\ 
 &=  \frac{\sum_{i=1}^p\lambda_iy_i^2}{\sum_{i=1}^py_i^2} 
 \leq \lambda_1 \frac{\sum_{i=1}^py_i^2}{\sum_{i=1}^py_i^2}
 = \lambda_1 
\end{align*}

Now we show that this is actually attained for $\symbf{x}=\symbf{e}_1$. Since eigenvectors are by convention length 1, we consider $\symbf{e}_1'\symbf{B}\symbf{e}_1$.

First, let $\symbf{c}_i$ be the unit vector with a 1 in the $i$-th position (and 0's everywhere else).

Expanding $\symbf{B}$ by the eigen decomposition, we have that

$$
\symbf{e}_1'\symbf{B}\symbf{e}_1=\symbf{e}_1'\symbf{P}\symbf{\Lambda}\symbf{P}'\symbf{e}_1.
$$

Since $B$ is symmetric, the eigenvectors can be chosen orthogonal.
We do so, which implies that $\symbf{e}_i'P=\symbf{c}_i'$, where $\symbf{c}_i$ is the unit vector with a 1 in the $i$-th position (and 0 everwhere else).
Consequently,

$$
\symbf{e}_i'\symbf{P}\symbf{\Lambda}\symbf{P}'\symbf{e}_i
= \symbf{c}_i'\symbf{\Lambda}\symbf{c}_i=\lambda_i,
$$

and so $\symbf{e}_1'\symbf{P}\symbf{\Lambda}\symbf{P}'\symbf{e}_1=\lambda_1$.

Now, we consider the case where $\symbf{x}$ is orthogonal to $\symbf{e}_1, \symbf{e}_2, \ldots, \symbf{e}_{i-1}$.

Each component in the vector $\symbf{y}$ is the dot product of $x$ and an eigenvector $\symbf{e}_i$.
Since we choose $x$ orthogonal to the first $i-1$ eigenvectors, the first $i-1$ entries of $\symbf{y}$ are zero.

Returning to considering the quadratic form, we have that

\begin{align}
\frac{\symbf{x}'\symbf{B}\symbf{x}}{\symbf{x}'\symbf{x}} 
&= \frac{\sum_{j=1}^p\lambda_jy_j^2}{\sum_{j=1}^py_j^2} \\
&= \frac{\sum_{j=i}^p\lambda_jy_j^2}{\sum_{j=i}^py_j^2} 
 \leq \lambda_i \frac{\sum_{j=i}^py_j^2}{\sum_{j=i}^py_j^2} 
 = \lambda_i
\end{align}

Using the same argument as before, we can show that this is actually attained for $\symbf{x}=\symbf{e}_i$.

### Connection with principal components

- Theorem 1 has already done the heavy lifting, as we can simply set $\symbf{B}=\symbf{\Sigma}$.
  - To see this, consider that $\frac{\symbf{x}'\symbf{\Sigma}\symbf{x}}{\symbf{x}'\symbf{x}}=\frac{\symbf{x}'\symbf{\Sigma}\symbf{x}}{||\symbf{x}||^2} = (\symbf{x}/||\symbf{x}||)'\symbf{\Sigma}(\symbf{x}/||\symbf{x}||)$, so implicitly the constraint $||\symbf{x}||=1$ is already present.
- All that's left to do is place Theorem 1 in a probabilistic context, and relate it to the variance and covariance of linear combinations of a random vector.

## Theorem 2: Selecting principal components

Let $\symbf{X}:\symbf{p}\times 1$ be the random vector with variance-covariance matirx $\symbf{\Sigma}:p\times p$, which has the $i$-th largest eigenvalue as $\lambda_i$ and associated eigenvector $\symbf{e}_i$.

Then the $i$-th principal component (defined as before) is given by 

$$
Y_i = \symbf{e}_i'\symbf{X},
$$

implying that

\begin{align*}
\mathrm{Var}[Y_i] &= \lambda_i \;\forall \; i \in \{1,2, \ldots, p\}, \; \mathrm{and} \\
\mathrm{Cov}[Y_i, Y_j] &= 0 \; \mathrm{for} \; i\neq j.
\end{align*}

### Proof of Theorem 2

From Theorem 1, we know that 

$$
\max_{\symbf{x}} \frac{\symbf{x}'\symbf{\Sigma}\symbf{x}}{\symbf{x}'\symbf{x}} = \lambda_1,
$$

which we achieve by setting $\symbf{x}=\symbf{e}_1$.

Since $||\symbf{e}_1||=1$, $\frac{\symbf{e}_1'\symbf{\Sigma}\symbf{e}_1}{\symbf{e}_1'\symbf{e}_1}=\symbf{e}_1'\symbf{\Sigma}\symbf{e}_1$, which is equal to $\mathrm{Var}[\symbf{a}'\mathrm{X}]=\mathrm{Var}[Y_1]$ if we set $\symbf{a}=\symbf{e}_1$. This implies that $\symbf{a}=\symbf{e}_1$ maximises the variance, which is $\lambda_1$.

Analagous reasoning shows that $\mathrm{Var}[\symbf{a}_i'\symbf{X}]$ has a maximum value $\lambda_i$ for $i\in\{2,3,\ldots,p\}$ that is attained when we set $\symbf{a}_i=\symbf{e}_i$, under the restriction $\symbf{a}_i$ to be orthogonal to $\symbf{a}_1, \symbf{a}_2, \ldots, \symbf{a}_{i-1}$.

For $i\neq j$, $\mathrm{Cov}[\symbf{e}_i'\symbf{X},\symbf{e}_j'\symbf{X}] = \symbf{e}_i'\symbf{\Sigma}\symbf{e}_j = 0$.

## A note on terminology

- Principal components (the $\symbf{Y}_i$'s) may also be referred to as "scores".
- The coefficient vectors $\symbf{a}_i$ may also be referred to as "loadings" (thinking algebraically) or as "principal axes"/"principal directions" (thinking geometrically).

## PCA as rotation

- PCA can be viewed as a rotation of the original variables to new axes.
- The new axes are the principal components, and the original variables are projected onto these new axes.
- The first principal component is the direction of maximum variance, and the second principal component is orthogonal to the first and captures the next highest variance, and so on.
- This rotation helps to simplify the data structure and highlight the most important features of the data.

### First principal component

- The first PC can be visualised as an axis along the vector $\symbf{a} = [1/\sqrt{2}, 1/\sqrt{2}]$, with the projection of the point $[1.2, 0.3]$ onto this axis being the new variable:

```{r}
set.seed(2106)
plot_tbl <- tibble::tibble(
  height = rnorm(100, 160, 10)
) |>
  dplyr::mutate(
    weight = height / 2.5 + rnorm(100, 0, 3.5 * height / 160)
  ) |>
  dplyr::mutate(
    across(everything(), function(x) (x - mean(x)) / sd(x))
  ) |>
  dplyr::arrange(height)
proj_list <- list(c(1 / sqrt(2), 1 / sqrt(2)), c(1 / sqrt(2), -1 / sqrt(2)))[1]
plot_tbl_ind <- tibble::tibble(
  height = 1.2,
  weight = 0.3
)
plot_tbl_proj_ind <- proj_list |>
  purrr::map_df(~ mutate_proj(plot_tbl_ind, .x))

p <- ggplot(
  data = plot_tbl
) +
  geom_vline(
    xintercept = 0, col = "gray50", linetype = "solid"
  ) +
  geom_hline(
    yintercept = 0, col = "gray50", linetype = "solid"
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(
    x = "Height",
    y = "Weight"
  ) +
  geom_point(aes(x = height, y = weight), alpha = 0.1, col = "dodgerblue") +
  geom_point(
    data = plot_tbl_proj_ind,
    aes(x = height, y = weight), col = "mediumseagreen"
  ) +
  geom_segment(
    data = plot_tbl_proj_ind,
    aes(x = height, y = weight, xend = proj_x, yend = proj_y, group = proj),
    col = "gray25", arrow = arrow(length = unit(0.3, "cm")), linetype = "solid"
  ) +
  geom_abline(
    data = plot_tbl_proj_ind |>
      dplyr::mutate(slope = a2 / a1) ,
    aes(intercept = 0, slope = slope),
    col = "purple"
  ) +
  coord_cartesian(xlim = c(-2.5, 2.5), ylim = c(-2.5, 2.5)) +
  ggrepel::geom_text_repel(
    data = plot_tbl_proj_ind |>
      dplyr::mutate(a1 = a1 |> signif(2), a2 = a2 |> signif(2)),
    aes(x = a1, y = a2, label = proj),
    nudge_x = 0.1, nudge_y = 0.1
  )
path_p_height_weight_ex_1 <- projr::projr_path_get("cache", "fig", "p-height-weight-ex-1.png")
cowplot::save_plot(
  plot = p,
  filename = path_p_height_weight_ex_1,
  units = "cm",
  base_height = 8,
  base_width = 8
)
```

```{r}
#| results: asis
#| echo: false
knitr::include_graphics(path_p_height_weight_ex_1)
```

- If we set $\symbf{a} = [1, 0]$, then the "new" variable is the height.
- If we set $\symbf{a} = [0, 1]$, then the "new" variable is the weight.

- The point of these last two is that original coordinates are, in a sense, themselves projections.
- So, if we consider $[1, 0]$ and $[0, 1]$ as axes, then that helps to see why we can view $[1/\sqrt{2}, 1/\sqrt{2}]$ as a new axis.

### Second principal component

```{r}
set.seed(2106)
plot_tbl <- tibble::tibble(
  height = rnorm(100, 160, 10)
) |>
  dplyr::mutate(
    weight = height / 2.5 + rnorm(100, 0, 3.5 * height / 160)
  ) |>
  dplyr::mutate(
    across(everything(), function(x) (x - mean(x)) / sd(x))
  ) |>
  dplyr::arrange(height)
proj_list <- list(c(1 / sqrt(2), 1 / sqrt(2)), c(1 / sqrt(2), -1 / sqrt(2)))
plot_tbl_ind <- tibble::tibble(
  height = 1.2,
  weight = 0.3
)
plot_tbl_proj_ind <- proj_list |>
  purrr::map_df(~ mutate_proj(plot_tbl_ind, .x))

p <- ggplot(
  data = plot_tbl
) +
  geom_vline(
    xintercept = 0, col = "gray50", linetype = "solid"
  ) +
  geom_hline(
    yintercept = 0, col = "gray50", linetype = "solid"
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(
    x = "Height",
    y = "Weight"
  ) +
  geom_point(aes(x = height, y = weight), alpha = 0.1, col = "dodgerblue") +
  geom_point(
    data = plot_tbl_proj_ind,
    aes(x = height, y = weight), col = "mediumseagreen"
  ) +
  geom_segment(
    data = plot_tbl_proj_ind,
    aes(x = height, y = weight, xend = proj_x, yend = proj_y, group = proj),
    col = "gray25", arrow = arrow(length = unit(0.3, "cm")), linetype = "solid"
  ) +
  geom_abline(
    data = plot_tbl_proj_ind |>
      dplyr::mutate(slope = a2 / a1) ,
    aes(intercept = 0, slope = slope),
    col = "purple"
  ) +
  coord_cartesian(xlim = c(-2.5, 2.5), ylim = c(-2.5, 2.5)) +
  ggrepel::geom_text_repel(
    data = plot_tbl_proj_ind |>
      dplyr::mutate(a1 = a1 |> signif(2), a2 = a2 |> signif(2)),
    aes(x = a1, y = a2, label = proj),
    nudge_x = 0.1, nudge_y = 0.1
  )
path_p_height_weight_ex_2 <- projr::projr_path_get("cache", "fig", "p-height-weight-ex-2.png")
cowplot::save_plot(
  plot = p,
  filename = path_p_height_weight_ex_2,
  units = "cm",
  base_height = 8,
  base_width = 8
)
```

```{r}
#| results: asis
#| echo: false
knitr::include_graphics(path_p_height_weight_ex_2)
```
