# Theorem three: Total variance {.smaller}

Let $\symbf{X}' = [X_1, X_2, \ldots, X_p]'$ be a random vector with covariance matrix $\symbf{\Sigma}$ that has eigenvalue-eigenvector pairs $(\lambda_1, \symbf{e}_1), \ldots, (\lambda_p, \symbf{e}_p)$, ordered such that $\lambda_1 \geq \lambda_2 \geq \ldots \geq \lambda_p \geq 0$. Define the principal components $Y_i = \symbf{e}'_i\symbf{X}$ for $i = 1, \ldots, p$. Then, the sum of variances of the original variables is equal to the sum of the eigenvalues of $\symbf{\Sigma}$, which is also equal to the sum of variances of the principal components:

$$
\sum_{i=1}^{p} \sigma_{ii} = \sum_{i=1}^{p} \text{Var}(X_i) = \sum_{i=1}^{p} \lambda_i = \sum_{i=1}^{p} \text{Var}(Y_i).
$$

# Proof of theorem three {.smaller}

The trace of the covariance matrix $\symbf{\Sigma}$, which is the sum of the variances of the original variables.

Since $\symbf{\Sigma}$ can be decomposed as $\symbf{\Sigma} = \symbf{PDP}'$, where $\symbf{D}$ is the diagonal matrix of eigenvalues and $\symbf{P}$ is the matrix of eigenvectors, we have:

$$
\text{trace}(\symbf{\Sigma}) = \text{trace}(\symbf{PDP}') = \text{trace}(\symbf{D}) = \sum_{i=1}^{p} \lambda_i = \sum_{i=1}^{p} \text{Var}(Y_i).
$$

# Contribution per component

Theorem 2 showed that

$$
\text{trace}(\symbf{\Sigma}) = \sum \lambda_i = \sum \text{Var}(Y_i).
$$

This implies that the proportion of the total variance due to the $k^{th}$ principal component is

$$
\frac{\lambda_k}{\sum \lambda_i },\quad k=1,...p,
$$

and (due to being uncorrelated) the proportion of variance accounted for by the first $r$ principal components is given by

$$
\frac{\sum_{i=1}^r \lambda_i }{\sum \lambda_i }.
$$

# The scree plot

:::: {.columns}

::: {.column width="50%"}

- The scree plot is a graphical representation of the proportion of variance accounted for by each principal component.
- It plots $(i, \lambda_i/\sum_{j=1}^{p} \lambda_j)$ for $i = 1, \ldots, p$.
- Typically, we look for the "elbow" in the plot, which indicates the number of principal components to retain.
- Beyond this point, each additional principal component captures little additional variation.

:::

::: {.column width="50%"}

```{r}
plot_tbl <- tibble::tibble(
  pc = 1:7,
  prop_var = eig_obj$values / sum(eig_obj$values)
)
plot_tbl_point <- purrr::map_df(
  seq(5.5, 5.502, length.out = 9), function(x) {
    plot_tbl |>
      dplyr::filter(pc == 2) |>
      dplyr::mutate(size_point = x)
  }
)
p <-  ggplot(
  data = plot_tbl
  ) +
    geom_line(
      aes(x = pc, y = prop_var),
      stat = "identity",
      colour = "dodgerblue"
    ) +
    geom_point(
      data = plot_tbl_point |> dplyr::slice(5:9),
      aes(x = pc, y = prop_var, size = size_point),
      colour = "mediumseagreen",
      alpha = 0.9,
      shape = 1,
      show.legend = FALSE
    ) +
    geom_point(
      aes(x = pc, y = prop_var),
      colour = "dodgerblue",
      alpha = 0.9,
      size = 1.5
    ) +
    labs(
      x = "Principal component index",
      y = "Proportion of variance"
    ) +
    cowplot::theme_cowplot() +
    cowplot::background_grid(major = "xy") +
    theme(
      panel.background = element_rect(fill = "white"),
      plot.background = element_rect(fill = "white")
    ) +
    scale_x_continuous(
      breaks = 1:7
    )
path_p_scree <- projr::projr_path_get("cache", "fig", "p-scree.pdf")
cowplot::save_plot(
  plot = p,
  filename = path_p_scree,
  units = "cm",
  base_height = 7,
  base_width = 8 * (7/6)
)
```

```{r}
#| results: asis
knitr::include_graphics(path_p_scree)
```

:::

::::

# Theorem 4: Correlation with original variables

For $Y_i$ and $X_k$ the $i$-th principal component and the $k$-th original variable, respectively, we have that

$$
\rho_{Y_i,X_k} = \frac{\sqrt{\lambda_i}e_{ik}}{\sqrt{\sigma_{kk}}}.
$$


# Proof of theorem 3

Let $\symbf{a}_k'$ equal 1 in position $k$ and 0 otherwise, so that $X_k = \symbf{f}_k'\symbf{X}$. We note that

$$
\text{Cov}(X_k,Y_i) = \text{Cov}(\symbf{a}_k'\symbf{X},\symbf{e}_i'\symbf{X})=\symbf{a}_k'\symbf{\Sigma}\symbf{e}_i=\symbf{a}_k'\symbf{PDP'}\symbf{e}_i=\lambda_i\symbf{a}_k'\symbf{e}_i=\lambda_ie_{ik}.
$$

We know that $\text{Var}(Y_i)= \lambda_i$ and $\text{Var}(X_k)=\sigma_{kk}$, so

$$
\rho_{Y_i,X_k}=\frac{\text{Cov}(Y_i,X_k)}{\sqrt{\text{Var}(Y_i)}\sqrt{\text{Var}
(X_k)}}= \frac{\sqrt{\lambda_i}e_{ik}}{\sqrt{\sigma_{kk}}}
$$

# Example {.smaller}

::::{.columns}

:::{.column width="70%"}

- From our previous example where we had two sets of correlated variables, we obtain the following correlation matrix: 

```{r}
#| results: asis
P <- eig_obj$vectors
D <- diag(eig_obj$values)
var_mat <- diag(diag(vcov_mat))
numerator <- P %*% sqrt(D)
corr_mat <- var_mat %*% numerator
rownames(corr_mat) <- paste0("V", 1:7)
colnames(corr_mat) <- paste0("PC", 1:7)
corr_mat |> signif(2) |> knitr::kable()
```

:::

:::{.column width="30%"}

- As all variables have unit variance, the only things that matter are the eigenvalue and the eigenvector.
  - High loading and high eigenvalue $\rightarrow$ high correlation.
- Higher variance (hypothetically) decorrelates PCs and original variables.

:::

::::

# Constant-density contours I {.smaller}

By the definition of the MVN PDF, the density of $X$ is constant on the ellipsoid defined by the equation

$$
(\symbf{x} - \mu)'\Sigma^{-1}(\symbf{x} - \mu) = c^2.
$$

The ellipsoid's shape, size and orientation are determined by the eigendecomposition of $\symbf{\Sigma}$.

WLOG, we assume $\symbf{\mu} = \symbf{0}$. Then, the ellipsoid is defined by

$$
c^2 = x'\Sigma^{-1}x = \frac{1}{\lambda_1}(\symbf{e}_1'\symbf{x})^2 + \frac{1}{\lambda_2}(\symbf{e}_2'\symbf{x})^2 + \ldots + \frac{1}{\lambda_p}(\symbf{e}_p'\symbf{x})^2
$$

after multiplying out.

# Constant-density contours (cont.) {.smaller}

The axes of the ellipse are then given by

$$
\pm c\sqrt{\lambda_i}\symbf{e}_i, \quad i = 1,2,\ldots,p.
$$

The eigenvectors therefore define the direction and the eigenvalues the length of the axes of the ellipsoid.

Furthermore, if we set $y_i = \symbf{e}_i'\symbf{x}$ (the $i$-th principal componet for $\symbf{x}$), we have

$$
c^2 = \frac{1}{\lambda_1}y_1^2 + \ldots + \frac{1}{\lambda_p}y_p^2.
$$

The equation is therefore satisfied by the principal components of $\symbf{x}$.

# Variance standardisation

- Before applying PCA, one can standardise the variances to have unit variance.
- This may be appropriate when:
  - the variances have different units (e.g. weight in kg vs annual income in rands), or
  - the variances have different scales (e.g. some genes are extremely rare whereas others are abundantly expressed), or
  - we wish to eliminate any effect of differences in variance on downstream analyses, e.g. when applying penalised regression.
- Similar comments go for logging the data, or applying other transformations.
- All results go through in exactly the same way, with adjustments made in interpreting exactly what variables the principal components capture.

# Standardisation example

::::{.columns}

:::{.column width="50%"}

We have the following covariance matrix:

```{r}
cov_mat <- matrix(
  c(1, 4, 4, 100), nrow = 2
)
cov_mat
```

which produces the following eigendecomposition:

```{r}
eigen(cov_mat) |> lapply(function(x) signif(x, 2))
```

:::

:::{.column width="50%"}

The associated correlation matrix:

```{r}
diag_mat_sd_inv <- diag(diag(cov_mat), nrow = nrow(cov_mat)) |> sqrt() |> solve()
rho_mat <- diag_mat_sd_inv %*% cov_mat %*% diag_mat_sd_inv
rho_mat 
```

has the following eigendecomposition:

```{r}
eigen(rho_mat) |> lapply(function(x) signif(x, 2))
```

The eigenvectors are in very different directions.

:::

::::

# Sample principal components {.smaller}

::::{.columns}

:::{.column width="50%"}

In practice, we do not know $\symbf{\Sigma}$ and must estimate it by the sample covariance matrix $\symbf{S}=n^{-1}\symbf{X}'\symbf{X}$ (assuming $\symbf{X}$ is mean-centred; adjust otherwise).

For example, sample covariance matrix of the height and weight data is:

```{r}
#| echo: true
#| results: asis
cov_mat <- hw_tbl |>
  as.matrix() |>
  cov()
cov_mat |> signif(2) |> knitr::kable()
```

:::

:::{.column width="50%"}

- The eigen decomposition is then:

```{r}
#| echo: true
eigen(cov_mat) |>
  lapply(function(x) signif(x, 2))
```

So, the first principal component is given by projections onto $[1, 1]$.

:::

::::


# Sample principal components (cont.) {.smaller}

If $\symbf{X} \sim \mathcal{N}(\symbf{0}, \symbf{\Sigma})$, then we have that that

$$
Y_i \sim \mathcal{N}(0, \lambda_i) \quad \text{and} \quad \symbf{Y} \sim \mathcal{N}_p(\symbf{0}, \symbf{\Lambda}).
$$

In the height and weight example then, we estimate this to be

$$
\symbf{Y} \sim \mathcal{N}_p(\symbf{0}, \begin{bmatrix} 1.71 & 0 \\ 0 & 0.31 \end{bmatrix}).
$$

# Approximating a matrix $\symbf{X}$

Suppose we have a matrix $\symbf{X}$ of size $n \times p$ with rank $t$ and we wish to approximate it with a matrix $\hat{\symbf{X}}$ of rank $s < t$.

In other words, we wish to find a matrix

$$
\hat{\symbf{X}} = \symbf{A}\symbf{B},
$$

where $\symbf{A}$ is of size $n \times s$ and $\symbf{B}$ is of size $s \times p$, such that

$$
\symbf{X} \approx \hat{\symbf{X}}.
$$

We want a rank $s$ approximation to the matrix $\symbf{X}$.

# Approximation example 

In the height and weight dataset, we obtained a size variable (the first principal component) that was an equally-weighted combination of weight and height, as follows:

$$
\text{size} = \frac{1}{\sqrt{2}}\text{height} + \frac{1}{\sqrt{2}}\text{weight}. = \begin{pmatrix} \text{height} && \text{weight} \end{pmatrix} \begin{pmatrix} \frac{1}{\sqrt{2}} \\ \frac{1}{\sqrt{2}} \end{pmatrix}.
$$

We'll show that can approximate the original height and weight variables using the size variable:

$$
[\text{height}, \text{weight}] \approx \text{size} \begin{pmatrix} \frac{1}{\sqrt{2}} & \frac{1}{\sqrt{2}} \end{pmatrix}  = \begin{pmatrix} 1/\sqrt{2} \text{size} & 1/\sqrt{2} \text{size} \end{pmatrix}.
$$

We therefore have a "rule" for mapping size onto weight and height.

# Approximation example (cont.) {.smaller}

Suppose that a given observation has height 0.7 and weight 0.5.

This implies a first principal component (size) of

$$
\text{size} = \frac{1}{\sqrt{2}}\times 0.7 + \frac{1}{\sqrt{2}}\times 0.5 = 0.85.
$$

We can approximate the height and weight as follows:

$$
\begin{pmatrix} \hat{\text{height}} & \hat{\text{weight}} \end{pmatrix} \approx \begin{pmatrix} 1/\sqrt{2} \times 0.85 & 1/\sqrt{2} \times 0.85 \end{pmatrix} = \begin{pmatrix} 0.6 & 0.6 \end{pmatrix}.
$$

Considering the standard deviation is (by construction) 1, the error (0.1 in both cases) is small.

If weight and height were very different, the error would be larger (why?).

# Why are about approximation?

Whether or not PCA is a good approximation matters because

a) we may actually want to reconstitute the original data after compression (e.g. image compression),
b) we want to know we're not losing too much information and
c) it helps provide some intuition about what we're doing.

# Minimising the least squares error of the approximation

We wish to find a matrix $\hat{\symbf{X}}$ of rank $s$ that that minimises the least squares error:

$$
\text{min}_{\hat{\symbf{X}}} \sum_{i=1}^{n} \sum_{j=1}^{p} (x_{ij} - \hat{x}_{ij})^2 = \text{min}_{\hat{\symbf{X}}} \text{trace}( (\symbf{X} - \hat{\symbf{X}})(\symbf{X} - \hat{\symbf{X}})').
$$

We are going to

1. prove that SVD can be used to find $\hat{\symbf{X}}$, and
2. that the PCA and SVD are equivalent in this context.

# Theorem 5a: SVD approximation

Let $\symbf{X}$ be a matrix of size $n \times p$ with rank $t$ and SVD $\symbf{X} = \symbf{U}\symbf{D}\symbf{V}'$. Then the best rank $s$ approximation to $\symbf{X}$ is given by

$$
\hat{\symbf{X}} = \symbf{U}\symbf{D}\symbf{J}_s\symbf{V}',
$$

where $\symbf{J}_s$ is the matrix of size $p \times p$ with the first $s$ diagonal elements equal to 1 and the rest equal to 0.

This is equivalent to

$$
\hat{\symbf{X}} = \sum_{i=1}^{s} d_i\symbf{u}_i\symbf{v}_i',
$$

where $\symbf{u}_i$ and $\symbf{v}_i$ are the left- and right-singular vectors and $d_i$ is the $i$-th singular value of $\symbf{X}$.

# Proof of Theorem 5a {.smaller} 

\begin{align*}
& \text{We use } UU^* = I_m \text{ and } VV^* = I_k \text{ to write the sum of squares as} \\
& \text{tr}[(A - B)(A - B)'] = \text{tr}[UU'(A - B)VV'(A - B)'] \text{ (by orthonormality)}, \\
& = \text{tr}[U'(A - B)VV'(A - B)U] \text{ (by cyclic trace property)}, \\
& = \text{tr}[U'(A - B)V(U'(A - B)V)'] \text{ (as }\mathrm{tr}(AB)=B'A'\text{)},\\
& = \text{tr}[(U'AV-U'BV)(U'AV-U'BV)^*] \text{ (multiplying in)}. \\
& \text{As } U'AV=U'UDV'V=D \text{ and letting } C = U'BV, \text{ from above we have that} \\
& \text{tr}[(A - B)(A - B)'] = \text{tr}[(D - C)(D - C)'] \\
& = \sum_{i,j=1}^{m} (d_{ij} - c_{ij})^2, \\
& = \sum_{i=1}^{m} (d_i - c_{ii})^2 + \sum_{i\neq j} c_{ij}^2 \text{ (as D is diagonal).}
\end{align*}

# Proof of Theorem 5a (cont.) {.smaller} 

The quantity $\sum_{i=1}^{m} (d_i - c_{ii})^2 + \sum_{i\neq j} c_{ij}^2$ is minimised when 

1. $c_{ij} = 0$ for $i \neq j$, 
2. $c_{ii} = d_i$ for $i \leq s$, and
3. $c_{ii}=0$ otherwise.

We choose the $s$ largest singular values as these reduce the error terms along the diagonals the most.
We cannot choose more than $s$ as we $\symbf{U}'\symbf{B}\symbf{V}$ is diagonal, meaning the number of non-zero diagonal elements is its rank and its rank must be $s$ or less.

This implies that 

$\symbf{U}'\symbf{B}\symbf{V} = \symbf{D}\symbf{J}_s \Rightarrow \symbf{B} = \symbf{U}\symbf{D}\symbf{J}_s\symbf{V}'.$

# Theorem 5b: PCA leads to same approximation as the SVD {.smaller}

The PCA approximation to $\symbf{X}$ is given by $\tilde{\symbf{X}}=\symbf{Y}\symbf{J}_k\symbf{P}'$, where $\symbf{J}_k$ is the matrix of size $p \times p$ with the first $k$ diagonal elements equal to 1 and the rest equal to 0.

This is the same as the SVD approximation, i.e.

$$
\hat{\symbf{X}} = \symbf{U}\symbf{D}\symbf{J}_k\symbf{V}' = \tilde{\symbf{X}} = \symbf{Y}\symbf{J}_k\symbf{P}'.
$$

# Theorom 5b: proof {.smaller}

The principal components are given by

$$
\symbf{Y} = \symbf{X}\symbf{P}.
$$

The first $k$ principal components are given by

$$
\symbf{Y}\symbf{J}_k = \symbf{X}\symbf{P}\symbf{J}_k,
$$

setting the last $p-k$ columns of $\symbf{P}$ to 0.

This implies that the approximation to $\symbf{X}$ is given by 

$$
\tilde{\symbf{X}} = \symbf{Y}\symbf{J}_k\symbf{P}'.
$$

# Theorom 5b: proof (cont.) {.smaller}

The columns of $\symbf{P}$ are the eigenvectors of $\symbf{X}'\symbf{X}$, which are the same as the right singular vectors of $\symbf{X}$, i.e.

$$
\symbf{P} = \symbf{V}.
$$

This implies that

$$
\tilde{\symbf{X}} = \symbf{Y}\symbf{J}_k\symbf{P} = \symbf{XV}\symbf{J}_k\symbf{V}' = \symbf{U}\symbf{D}\symbf{VV}'\symbf{J}_k\symbf{V}' = \symbf{U}\symbf{D}\symbf{J}_k\symbf{V}' = \hat{\symbf{X}}.
$$

# Example

:::: {.columns}

::: {.column width="57%"}

We read in data of measurements Painted Turtles [@jolicoeur_mosimann60]:

```{r}
#| eval: false
#| echo: true
if (!requireNamespace(
  "remotes", quietly = TRUE
  )) {
  install.packages("remotes")
}
"MiguelRodo/DataTidy23RodoHonsMult@2024" |>
  remotes::install_github()
```

:::

::: {.column width="43%"}

```{r}
#| echo: true
data(
  "data_tidy_turtle",
  package = "DataTidy23RodoHonsMult"
)
data_tidy_turtle
```

:::

::::

# Measurements highly correlated

::::{.columns}

:::{.column width="50%"}

```{r}
#| include: false
plot_tbl_turtle <- data_tidy_turtle |>
  tidyr::pivot_longer(
    length:width,
    names_to = "x_var",
    values_to = "x_val"
  )
p <- ggplot(
  plot_tbl_turtle,
  aes(x = x_val, y = height, col = gender)
) +
  geom_point(alpha = 0.75) +
  scale_colour_manual(
    values = c(male = "dodgerblue", female = "mediumseagreen")
   ) +
  facet_wrap(~x_var, scales = "free", ncol =1 ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme( 
    panel.background = element_rect(fill = "white"), 
    plot.background = element_rect(fill = "white")
  ) +
  labs(
    x = "Measurement",
    y = "Height"
  ) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  ) +
  theme(
    axis.ticks = element_blank(),
    axis.text = element_blank()
  ) +
  theme(
    strip.background = element_rect(fill = "white", colour = "gray25")
  )
path_p_turtle <- projr::projr_path_get("cache", "fig", "p-turtle.png")
cowplot::save_plot(
  plot = p,
  filename = path_p_turtle,
  units = "cm",
  base_height = 7,
  base_width = 5
)
```

```{r}
#| results: "asis"
knitr::include_graphics(path_p_turtle)
```

:::

:::{.column width="50%"}

- Due to the correlation, we may likely summarise the data very well using principal components.
- We'll use the `prcomp` function in R, and do it from first principles.

:::

::::

# Using the `prcomp` function I {.smaller}

::::{.columns}

:::{.column width="70%"}

```{r}
#| echo: true
pr_obj <- prcomp(
  ~log(length) + log(width) + log(height),
  data = data_tidy_turtle,
  retx = TRUE
)
pr_obj
```

:::

:::{.column width="30%"}

- The first principal component has a markedly higher standard deviation.
- It is weighted roughly evenly across the variables, indicating it is a general "size" variable.

:::

::::

# Using the `prcomp` function II {.smaller}

::::{.columns}

:::{.column width="70%"}

```{r}
#| echo: true
summary(pr_obj)
```

:::

:::{.column width="30%"}

- The `summary` function provides the per-component standard deviation and variance, as well as the cumulative variance.

:::

::::

# Using the `prcomp` function III {.smaller}

::::{.columns}

:::{.column width="70%"}

```{r}
#| echo: true
plot(pr_obj)
```

:::

:::{.column width="30%"}

- The `plot` command provides the scree plot.
- Clearly, unless we're specifically interested in non-size variables we should only retain the first principal component.

:::

::::

# Using the `prcomp` function IV {.smaller}

Here are the principal components:

```{r}
#| echo: true
pr_obj$x |> head() |> signif(2)
```

# From first principles I {.smaller}

Using the SVD, we can obtain the same results as the `prcomp` function.

Removing gender, logging and centering the data:

```{r}
#| echo: true
data_turtle_mat <- data_tidy_turtle |>
  dplyr::select(-gender) |>
  dplyr::mutate(across(everything(), log)) |>
  dplyr::mutate(across(everything(), function(x) x - mean(x))) |>
  as.matrix()
```

We then calculate the SVD:

```{r}
#| echo: true
svd_turtle <- svd(data_turtle_mat)
```

# From first principles II {.smaller}

The eigenvalues are the squares of the singular values:

```{r}
#| echo: true
svd_turtle$d^2 |> signif(3)
```

The right-singular vectors are the eigenvectors of the covariance matrix, and hence are the loading vectors, implying the principal components are given by:

```{r}
#| echo: true
(data_turtle_mat %*% svd_turtle$v) |> head() |> signif(2)
```

# Biplots

- Since each data point has many variables, graphical display is cumbersome.
- However, we may use PCA to provide an approximate graphical display of the data.
- It will show both:
  - Each observation, and
  - Each variable.
- Hence, "bi"plot.

# Definition of a biplot

Suppose we have a rank $s$ matrix $\hat{\symbf{X}}:n\times p$.
Then we can write

$$
\hat{\symbf{X}} = \symbf{A}\symbf{B},
$$

where $\symbf{A}$ is of size $n \times s$ and $\symbf{B}$ is of size $s \times p$.

The biplot is a plot of the rows of $\symbf{A}$ and the columns of $\symbf{B}$.

Typically, we plot the rows as points and the columns as vectors.

# A PCA biplot

The PCA approximation to a matrix $\symbf{X}$ is given by

$$
\symbf{Y}\symbf{J}_k\symbf{P}'=\symbf{Y}\symbf{J}_k\symbf{J}_k\symbf{P}'=\symbf{A}\symbf{B},
$$

where $\symbf{A}=\symbf{Y}\symbf{J}_k$ and $\symbf{B}=\symbf{J}_k\symbf{P}'$.

Typically, we set $k=2$ and ignore the all-zero columns/rows of $\symbf{A}$ and $\symbf{B}$.

The rows of $\symbf{Y}$ are the scores (representing the observations) and the rows of $\symbf{P}'$ are the loadings (representing the variables).

# Interpreting a biplot

- An observation has (approximately) an above-average value of a variable if it is close to the tip of the vector representing that variable.
- Two variables are (approximately) correlated if their vectors are close together.
  - They are (approximately) uncorrelated if they are orthogonal.

\centering

\includegraphics[width=0.65\textwidth]{_raw/img/2024/biplot.png}

# Resources

- Source code is on [GitHub](https://github.com/MiguelRodo/Slides23RodoHonsMult/blob/2024/SlidesHonsMultL7PCA.qmd).
  - Suppressed code (e.g. to create figures) is available there.

## References

